---
title: "LCA"
author: "L. Gebrekristos"
date: "1/15/2023"
output: html_document
---

## Purpose
Use latent class analysis (LCA) to create subgroups of sexual relationships based on 5 variables: knowledge of partner HIV status (noknowhiv), age disparity (agedisp), married or cohabiting (married), transactional sex (trans), and financial dependence (work)

```{r setup, include=FALSE,warning=FALSE,message=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=60),tidy=TRUE)

data<-read.csv("LCAdata")

# function to load needed packages
loadpackages <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg)) 
    install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}
```

## Needed Packages
```{r packages}
packages<-c("poLCA","dplyr","tidyLPA","sjPlot")
loadpackages(packages)
```

## LCA Models 
```{r lca models}
# define model: using 5 binary variables for LCA
f<-with(data, cbind(noknowhiv, agedisp, married, trans,work) ~1)
# models with 1 class to 4 classes
lcamodels<-function(numclass){
  poLCA(formula = f, data = data, nclass = numclass, maxiter=10000, verbose = F)
}

mod_1<-lcamodels(1)
mod_2<-lcamodels(2)
mod_3<-lcamodels(3)
mod_4<-lcamodels(4)
```

## Creating table with fit stats
```{r table}
# function to compare LCA models (LRT)
lrt <- function(lower,higher){
  # lower number of classes
  n <- lower$Nobs # number of observations 
  null_ll <- lower$llik # log-likelihood
  null_param <- lower$npar # number of parameters
  null_classes <- length(lower$P) # number of classes
  # higher number of classes
  alt_ll <- higher$llik # log-likelihood
  alt_param <- higher$npar # number of parameters
  alt_classes <- length(higher$P) # number of classes
  calc_lrt(n, null_ll, null_param, null_classes, alt_ll, alt_param, alt_classes)
}
# function for calculating entropy
entropy<-function(model){
  calc<-function (p) sum(-p*log(p))
error_prior<-calc(model$P) 
error_post<-mean(apply(model$posterior,1, calc),na.rm = TRUE)
round(((error_prior-error_post) / error_prior),2)
}
# creating dataframe for fit values
results <- data.frame(Modell=c("1"),
                      log_likelihood=lc1$llik,
                      df = lc1$resid.df,
                      BIC=lc1$bic,
                      ABIC=  (-2*lc1$llik) + ((log((lc1$N + 2)/24)) * lc1$npar),
                      CAIC = (-2*lc1$llik) + lc1$npar * (1 + log(lc1$N)), 
                      likelihood_ratio=lc1$Gsq)
# 1st column is number of classes
results$Modell<-as.integer(results$Modell)
results[1,1]<-c("1")
results[2,1]<-c("2")
results[3,1]<-c("3")
results[4,1]<-c("4")
# 2nd column is log likelihood
results[2,2]<-lc2$llik
results[3,2]<-lc3$llik
results[4,2]<-lc4$llik
# 3rd column is df 
results[2,3]<-lc2$resid.df
results[3,3]<-lc3$resid.df
results[4,3]<-lc4$resid.df
# 4th column is BIC
results[2,4]<-lc2$bic
results[3,4]<-lc3$bic
results[4,4]<-lc4$bic
# 5th column is adjusted BIC
results[2,5]<-(-2*lc2$llik) + ((log((lc2$N + 2)/24)) * lc2$npar)
results[3,5]<-(-2*lc3$llik) + ((log((lc3$N + 2)/24)) * lc3$npar)
results[4,5]<-(-2*lc4$llik) + ((log((lc4$N + 2)/24)) * lc4$npar)
# 6th column is consistent AIC
results[2,6]<- (-2*lc2$llik) + lc2$npar * (1 + log(lc2$N)) 
results[3,6]<- (-2*lc3$llik) + lc3$npar * (1 + log(lc3$N))
results[4,6]<- (-2*lc4$llik) + lc4$npar * (1 + log(lc4$N))
# 7th column is likelihood ratio
results[1,7]<-c("-") #class 1 is 0 because no lower class to compare to 
results[2,7]<-round(lc2$Gsq,2)
results[3,7]<-round(lc3$Gsq,2)
results[4,7]<-round(lc4$Gsq,2)
# 8th column is p-value for LRT
# p-value <.05 then higher class is better
lrtc1c2<-lrt(mod_1,mod_2)
lrtc2c3<-lrt(mod_2,mod_3)
lrtc3c4<-lrt(mod_3,mod_4)
results[1,8]<-c("-") #see comment for class 1 in likelihood ratio column
lrt1<-ifelse(lrtc1c2[4]<.0001,"<.0001",round(lrt12[4],2))
lrt2<-ifelse(lrtc2c3[4]<.0001,"<.0001",round(lrt23[4],2))
lrt3<-ifelse(lrtc3c4[4]<.0001,"<.0001",round(lrt34[4],2))
results[2,8]<-lrt1
results[3,8]<-lrt2
results[4,8]<-lrt3
# 9th column is entropy
results[1,9]<-c("-")
results[2,9]<-entropy(mod_2)
results[3,9]<-entropy(mod_3)
results[4,9]<-entropy(mod_4)
# creating column names for table
colnames(results)<-c("Classes","log-likelihood","resid. df","BIC","aBIC","cAIC","likelihood-ratio","p-value","Entropy")
lca_results<-results

sjPlot::tab_df(lca_results,title="LCA fit statistics (n=3335)")
```

## Best Fit model (Model 3): Variable distributions by classes
```{r bestfit}
####################
#THREE SOLUTION
####################
# reordering classes
probs.start<-lc3$probs.start
new.probs.start<-poLCA.reorder(probs.start, c(1,3,2))
lc<-poLCA(f, data, nclass=3, probs.start=new.probs.start, na.rm=TRUE,maxiter=10000)
sol <- data.frame(var=c("Class %",""))
sol[3:17,1]<-c("<b>Knowledge of Partner's HIV status</b>","No Knowledge","Knowledge","<b>Age Disparate Relationship</b>","5 or more years","Less than 5 years","<b>Current Marital Status</b>","Married","Not Married","<b>Transactional Sex</b>","Yes","No","<b>Worked in last 12 months</b>","No","Yes")
round(colMeans(lc$posterior)[1]*100,1)
sol[1,2]<-round(colMeans(lc$posterior)[1]*100,1) #class membership
sol[1,3]<-round(colMeans(lc$posterior)[2]*100,1)
sol[1,4]<-round(colMeans(lc$posterior)[3]*100,1)
sol[1,2:4] <- paste0(as.matrix(sol[1,2:4]), '%')
pro<-unlist(lc$probs)
proround<-round(pro*100,0)
# bold % that are >50
proround<-ifelse(proround>50,paste0("<b>",proround,"</b>%"),paste0(proround,"%"))
sol[c(2,3,6,9,12,15),2:4]<-c("")
sol[4,2:4]<-proround[1:3] #know
sol[5,2:4]<-proround[4:6]
sol[7,2:4]<-proround[7:9]#agedisp
sol[8,2:4]<-proround[10:12]
sol[10,2:4]<-proround[13:15]#married
sol[11,2:4]<-proround[16:18]
sol[13,2:4]<-proround[19:21]#trans
sol[14,2:4]<-proround[22:24]
sol[16,2:4]<-proround[25:27]#work
sol[17,2:4]<-proround[28:30]

colnames(sol)<-c(" ","Class 1","Class 2","Class 3")


tab_df(sol,title="3 class solution (n=3335)",col.header = names(sol))
```

